Description of raw CFEC data processing for NCEAS portfolio effects working group

### 2016-04-07 process_raw_data/process_raw_cfec_data.Rmd
Revised species and vessel info

---
title: "Description of raw CFEC data processing for NCEAS portfolio effects working group"
author: "Jordan Watson & Eric Ward"
date: "March 16, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(mgcv)
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r dataload, warnings=FALSE, message = FALSE}
load("Data/cfec_w_IDs.RData")
cfec = cfecnew
rm(cfecnew) # remove the old named object

@@

```{r delColumns}
isNull = c("F_FG_MGT","C_CFACT", "F_DISTRC", "F_ERRFLG", "F_AYK", "G_FED_AR","F_BTAREA",
"A_CITY", "A_STATE", "A_ZIP", "P_YEAR", "I_NAME", "I_NAMTYP", "F_POT_DR", "F_PRCENT", 
"C_ADFGST", "F_CTDATE", "F_LNDATE", "F_NUMBER", "F_VALUE", "F_POUNDS", "F_TNUMB",  
"F_BATCH", "F_PERIOD",  "F_EMBSEQ", "F_PROCCK", "F_WEEK",  "F_TTYPE",    
"F_ITEM", "F_FSHERY",  "G_GEAR", "G_AREA", "G_PRC_AR",  "P_YEAR", "P_PMTSEQ", 
"p_check", "P_ID", "P_FEETYP", "P_ADFG", "P_FEEFLG", "P_FSHBLE", "F_DISPOS", 
"F_MPGMNM", "D_OPER", "D_O_FSHY", "D_O_PNUM", "D_O_PCHK", "D_O_YSEQ", "F_SIZE", 
"D_O_YCHK", "adfg")
cfec = cfec[,-which(names(cfec)%in%isNull)]
rm(isNull)
```

@@

table(cfec$P_STATUS)
```

@@

# Lump all sharks together
cfec$specn[cfec$specn==691] = 689
cfec$specn[cfec$specn==689] = 689
cfec$specn[cfec$specn==692] = 689

# Lump flatfish (Shallow)
cfec$specn[cfec$specn==123] = 119 # rock
cfec$specn[cfec$specn==129] = 119 # starry flounder
cfec$specn[cfec$specn==127] = 119 # yellowfin
cfec$specn[cfec$specn==126] = 119 # butter

@@

# Lump rockfish, codes from here:
# Slope rockfish
cfec$specn[cfec$specn==141] = 144 # POP
cfec$specn[cfec$specn==136] = 144 # northern
cfec$specn[cfec$specn==151] = 144 # rougheye
cfec$specn[cfec$specn==166] = 144 # sharpchin
cfec$specn[cfec$specn==152] = 144 # shortraker
cfec$specn[cfec$specn==153] = 144 # redbanded
cfec$specn[cfec$specn==157] = 144 # silvergray
cfec$specn[cfec$specn==158] = 144 # redstripe
cfec$specn[cfec$specn==137] = 144 # bocaccio
cfec$specn[cfec$specn==144] = 144 # unid
# Demersal rockfish
cfec$specn[cfec$specn==138] = 168 # copper
cfec$specn[cfec$specn==145] = 168 # yelloweye
cfec$specn[cfec$specn==146] = 168 # Canary
cfec$specn[cfec$specn==147] = 168 # quillback
cfec$specn[cfec$specn==148] = 168 # tiger
cfec$specn[cfec$specn==149] = 168 # China
cfec$specn[cfec$specn==150] = 168 # rosethorn
# Pelagic rockfish
cfec$specn[cfec$specn==154] = 169 # dusky
cfec$specn[cfec$specn==155] = 169 # yellowtail
cfec$specn[cfec$specn==156] = 169 # widow
cfec$specn[cfec$specn==172] = 169 # dusky
cfec$specn[cfec$specn==173] = 169 # dark
cfec$specn[cfec$specn==142] = 169 # black
cfec$specn[cfec$specn==167] = 169 # blue

#  Whitefish and trout did not have text labels in the species database.
#  We created our own for easier reference but given this lack of detail
#  it seems reasonable to lump these species groups for the sake of clarity
# Whitefish
cfec$specn[cfec$specn==580] <- 580  # Whitefish
cfec$specn[cfec$specn==583] <- 580  # Least cisco
cfec$specn[cfec$specn==584] <- 580  # Arctic cisco
cfec$specn[cfec$specn==585] <- 580  # Bering cisco
cfec$specn[cfec$specn==589] <- 580  # Humpback whitefish
# Trout - convert all to the same number - we'll label as "trout" later
cfec$specn[cfec$specn==520] <- 520  # Arctic char
cfec$specn[cfec$specn==530] <- 520  # Dolly varden
cfec$specn[cfec$specn==531] <- 520  # Dolly varden anadromous
cfec$specn[cfec$specn==540] <- 520  # Steelhead
cfec$specn[cfec$specn==560] <- 520  # Cutthroat

@@

# Salmon roe (by species)
# We consolidate roe of different species into a single roe category
cfec$specn[cfec$specn %in% c(400:405)] <- 400
# In early years, there were separate specn for salmon roe and salmon. In later years,
# salmon roe got absorbed into the specn for salmon but given a F_DELIVR==14.
# We may want to change this at some point, but for now, if F_DELIVR==14, then change
# the specn to that for roe instead of the specn "salmon."
cfec$specn[(cfec$specn %in% c(410,420,430,440,450)) & cfec$F_DELIVR=="14"] <- 400

@@

#  Rename some of the groups
cfec$spec <- as.character(cfec$spec)
cfec$spec[cfec$specn==700] = "SKATE"
cfec$spec[cfec$specn==689] = "SHARK"
cfec$spec[cfec$specn==119] = "FLAT.shallow"
cfec$spec[cfec$specn==118] = "FLAT.deep"
cfec$spec[cfec$specn==144] = "ROCK.slope"
cfec$spec[cfec$specn==168] = "ROCK.demersal"
cfec$spec[cfec$specn==169] = "ROCK.pelagic"
cfec$spec[cfec$specn==520] = "TROUT"

# Lump king crabs
cfec$spec[cfec$spec%in%c("GKCR","BKCR","UKCR","RKCR")] = "KCRB" # king
# Lump tanner crabs
cfec$spec[cfec$spec%in%c("UTCR","BTCR","aTCGR","aTCTR")] = "TCRB" # tanner
# Lump shrimp - spot prawn, pink shrimp, coonstriped 
cfec$spec[cfec$spec%in%c("SPRW","PSHP","USRM","aSSHP","aCOOS")] = "USRM" # shrimp

@@

# drop records with no spp
cfec = cfec[-which(is.na(cfec$spec)),]
```

@@

@@

## Output data file for use on other projects
```{r}
save(cfec, file = "cfecCleaned.Rdata")
```

@@

>>>>

---
title: "Description of raw CFEC data processing for NCEAS portfolio effects working group"
author: "Jordan Watson & Eric Ward"
date: "March 16, 2016 (Update 4/7/2016)""
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
#library(mgcv)
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r dataload, warnings=FALSE, message = FALSE}
setwd("//nmfs.local/AKC-ABL/Users2/jordan.watson/Desktop/AFSC/GOA/AKFIN/Meeting")
load("Data/cfec_w_IDs.RData")
cfec = cfecnew
rm(cfecnew) # remove the old named object

@@

```{r delColumns}
isNull = c("F_FG_MGT","C_CFACT", "F_DISTRC", "F_ERRFLG", "F_AYK", "G_FED_AR","F_BTAREA",
"A_CITY", "A_STATE", "A_ZIP", "P_YEAR", "I_NAME", "I_NAMTYP", "F_POT_DR", "F_PRCENT", 
"C_ADFGST", "F_CTDATE", "F_LNDATE", "F_NUMBER", "F_VALUE", "F_POUNDS", "F_TNUMB",  
"F_BATCH", "F_PERIOD",  "F_EMBSEQ", "F_PROCCK", "F_WEEK",  "F_TTYPE",    
"F_ITEM", "F_FSHERY",  "G_GEAR", "G_AREA", "G_PRC_AR",  "P_YEAR", "P_PMTSEQ", 
"p_check", "P_ID", "P_FEETYP", "P_ADFG", "P_FEEFLG", "P_FSHBLE", 
"F_MPGMNM", "D_OPER", "D_O_FSHY", "D_O_PNUM", "D_O_PCHK", "D_O_YSEQ", "F_SIZE", 
"D_O_YCHK", "adfg")
cfec = cfec[,-which(names(cfec)%in%isNull)]
rm(isNull)
```

@@

#table(cfec$P_STATUS)
```

@@

# Lump all sharks together
cfec$specn[cfec$specn==689] = 689
cfec$specn[cfec$specn==690] = 689
cfec$specn[cfec$specn==691] = 689
cfec$specn[cfec$specn==692] = 689

# Lump flatfish (Shallow)
cfec$specn[cfec$specn==123] = 119 # rock
cfec$specn[cfec$specn==129] = 119 # starry flounder
cfec$specn[cfec$specn==127] = 119 # yellowfin
cfec$specn[cfec$specn==126] = 119 # butter
cfec$specn[cfec$specn==122] = 119 # flathead
cfec$specn[cfec$specn==125] = 119 # rex
cfec$specn[cfec$specn==128] = 119 # English
cfec$specn[cfec$specn==131] = 119 # petrale
cfec$specn[cfec$specn==132] = 119 # sand
cfec$specn[cfec$specn==133] = 119 # Alaska plaice
cfec$specn[cfec$specn==116] = 119 # Bering flounder
cfec$specn[cfec$specn==117] = 119 # Kamchatka flounder

@@

# Lump rockfish, codes from here:
# Slope rockfish
cfec$specn[cfec$specn==136] = 144 # northern
cfec$specn[cfec$specn==137] = 144 # bocaccio
cfec$specn[cfec$specn==141] = 144 # POP
cfec$specn[cfec$specn==143] = 144 # thornyhead
cfec$specn[cfec$specn==144] = 144 # unid
cfec$specn[cfec$specn==151] = 144 # rougheye
cfec$specn[cfec$specn==152] = 144 # shortraker
cfec$specn[cfec$specn==153] = 144 # redbanded
cfec$specn[cfec$specn==157] = 144 # silvergray
cfec$specn[cfec$specn==158] = 144 # redstripe
cfec$specn[cfec$specn==159] = 144 # darkblotched
cfec$specn[cfec$specn==166] = 144 # sharpchin
cfec$specn[cfec$specn==176] = 144 # harlequin
cfec$specn[cfec$specn==177] = 144 # blackgill
cfec$specn[cfec$specn==179] = 144 # pygmy
cfec$specn[cfec$specn==181] = 144 # shortbelly
cfec$specn[cfec$specn==182] = 144 # splitnose
cfec$specn[cfec$specn==183] = 144 # stripe tail
cfec$specn[cfec$specn==184] = 144 # vermillion
cfec$specn[cfec$specn==185] = 144 # aurora

# Demersal rockfish
cfec$specn[cfec$specn==135] = 168 # greenstripe
cfec$specn[cfec$specn==138] = 168 # copper
cfec$specn[cfec$specn==139] = 168 # unid
cfec$specn[cfec$specn==140] = 168 # red
cfec$specn[cfec$specn==145] = 168 # yelloweye
cfec$specn[cfec$specn==146] = 168 # Canary
cfec$specn[cfec$specn==147] = 168 # quillback
cfec$specn[cfec$specn==148] = 168 # tiger
cfec$specn[cfec$specn==149] = 168 # China
cfec$specn[cfec$specn==150] = 168 # rosethorn
cfec$specn[cfec$specn==171] = 168 # shortraker/rougheye
cfec$specn[cfec$specn==175] = 168 # yellowmouth
cfec$specn[cfec$specn==178] = 168 # chilipepper

# Pelagic rockfish
cfec$specn[cfec$specn==142] = 169 # black
cfec$specn[cfec$specn==154] = 169 # dusky
cfec$specn[cfec$specn==155] = 169 # yellowtail
cfec$specn[cfec$specn==156] = 169 # widow
cfec$specn[cfec$specn==167] = 169 # blue
cfec$specn[cfec$specn==172] = 169 # dusky
cfec$specn[cfec$specn==173] = 169 # dark

# Group sculpins
cfec$specn[cfec$specn %in% c(160:165)] = 160

# Group greenlings (not AMCK or lingcod)
cfec$specn[cfec$specn %in% c(190,191,192,194)] = 190

# Group all osmerids together
cfec$specn[cfec$specn %in% c(510:516)] <- 510

# Group all char
cfec$specn[cfec$specn %in% c(520:532)] <- 520  # char

# Group all trout
cfec$specn[cfec$specn%in%c(540:562)] <- 540  # trout

#  Whitefish and trout did not have text labels in the species database.
#  We created our own for easier reference but given this lack of detail
#  it seems reasonable to lump these species groups for the sake of clarity
# Whitefish
cfec$specn[cfec$specn==580] <- 580  # Whitefish
cfec$specn[cfec$specn==583] <- 580  # Least cisco
cfec$specn[cfec$specn==584] <- 580  # Arctic cisco
cfec$specn[cfec$specn==585] <- 580  # Bering cisco
cfec$specn[cfec$specn==589] <- 580  # Humpback whitefish

@@

# Chinook have two species codes. Consolidate.
cfec$specn[cfec$specn==411] <- 410

# Salmon roe (by species) get lumped into a single salmon roe category. This was a topic of much discussion but ultimately, since much of the roe is unidentified, any other assignments would end up omitting a large percentage of the roe landings. 
cfec$specn[cfec$specn %in% c(401:405)] <- 400
cfec$specn[cfec$specn%in%c(410:450) & cfec$F_DELIVR==14] <- 400

# Herring are listed under 5 different species codes but 231:235 only account for a few thousand bucks. They are mostly bycatch or non-commercial.
# We'll lump them into the sac-roe species code because they will have no effect on the numbers. But this way we are consolidating codes instead
# of making a bunch of extra ones.
cfec$specn[cfec$specn %in% c(231:235)] <- 23043

# Most herring landings occur under species 230, with different F_DELIVR codes
# Create a species code for the sac-roe fishery
cfec$specn[cfec$specn==230 & cfec$F_DELIVR==43] <- 23043
# Create a species code for the bait/food fishery
cfec$specn[cfec$specn==230 & cfec$F_DELIVR==44] <- 23044
# Create a species code for the pound fishery
cfec$specn[cfec$specn==230 & cfec$F_DELIVR==45] <- 23045
# There may be 1 or 2 records with other codes, and which probably have no dollar value.
# Lump them into the sac roe fishery to avoid having extra codes
cfec$specn[cfec$specn==230] <- 23043

# Lump all king crabs
cfec$specn[cfec$specn %in% c(920:925)] <- 920

# Lump all king crabs
cfec$specn[cfec$specn %in% c(930:934)] <- 930

# Lump all remaining crabs (not Dungeness)
cfec$specn[cfec$specn %in% c(900,940:953)] <- 940

# Lump all shrimp
cfec$specn[cfec$specn %in% c(960:965)] <- 960

# Lump non-geoduck and non-razor clams together
cfec$specn[cfec$specn %in% c(810,812,820,825,840,842)] <- 810

# Lump weathervane and pink scallops
cfec$specn[cfec$specn %in% c(850:851)] <- 850

# Lump urchins
cfec$specn[cfec$specn %in% c(892:893,896)] <- 892

# Lump sea cucumbers
cfec$specn[cfec$specn %in% c(894:895)] <- 895

# Create an "OTHER" category that includes misc freshwater and a bunch of random trawl caught spec
cfec$specn[cfec$specn %in% c(100,101,112,170,180,190,206:220,250,260,310:380,490,500,520,540,570,580,590,600:683,714:800,880,890,855,899,998,999)] <- 206

@@

#  Rename some of the groups
cfec$spec <- as.character(cfec$spec)

#  Rename some of the groups
cfec$spec <- as.character(cfec$spec)
cfec$spec[cfec$specn==118] = "FLAT.deep"
cfec$spec[cfec$specn==119] = "FLAT.shallow"
cfec$spec[cfec$specn==144] = "ROCK.slope"
cfec$spec[cfec$specn==168] = "ROCK.demersal"
cfec$spec[cfec$specn==169] = "ROCK.pelagic"
cfec$spec[cfec$specn==206] = "OTHER"
cfec$spec[cfec$specn==400] = "aUROE"
cfec$spec[cfec$specn==689] = "SHARK"
cfec$spec[cfec$specn==700] = "SKATE"
cfec$spec[cfec$specn==810] = "OCLM"
cfec$spec[cfec$specn==850] = "SCAL"
cfec$spec[cfec$specn==892] = "URCH"
cfec$spec[cfec$specn==920] = "KCRB"
cfec$spec[cfec$specn==930] = "TCRB"
cfec$spec[cfec$specn==940] = "OCRB"
cfec$spec[cfec$specn==960] = "USRM"
cfec$spec[cfec$specn==23043] = "SAC"
cfec$spec[cfec$specn==23044] = "BAIT"
cfec$spec[cfec$specn==23045] = "POUND"

@@

# drop records with no spp
cfec = cfec[-which(is.na(cfec$spec)),]
rm(spec.tbl)
```

@@

cfec <- as.data.frame(cfec)   
```

Add vessel lengths to data
```{r}
#  Load the vessel characteristics data that were created using "Vessel_Characteristics.Rmd."
load("Data/vessel.RData")

#  For now, we only really need the vessel length data but different/additional fields can be added later by changing the select statement below. Note, I use "dplyr::select" instead of just "select" because if you have the plyr package loaded also, the select statement may get confused.

#  Convert fields to make them compatible with the vessel data.
cfec$cadfg2 <- as.numeric(cfec$cadfg)
cfec$year <- as.numeric(cfec$year)

#  I use a right join here instead of left join for continuity since "vessel" was the already loaded data frame which automatically puts it as the first argument of the piped command.
cfec <- dplyr::select(vessel,year,adfg,VLength) %>% right_join(cfec,by=c("adfg"="cadfg2","year"="year")) %>% rename(cadfg2=adfg)

#  Create an index value so that we can link back to the original (unscrambled data) if necessary.
cfec$INDEX <- 1:nrow(cfec)

@@

## Output data file for use on other projects
```{r}
save(cfec, file = "cfecCleaned2.Rdata")
```

@@

*specn.orig* The original version of the specn field before species were lumped into categories.

*spec* A text code created for each species. Some of these are the original AKFIN names and others were created by us. See the Species_Categories.xls file for descriptions.
